---
title: Aligned Multiple Factor Analysis
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aligned Multiple Factor Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE) && exists("params")) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
```

## Why aligned MFA?

Standard MFA assumes every block has exactly the same rows. `aligned_mfa()`
relaxes that assumption. You use it when blocks point back to the same latent
population, but each block observes only a subset of rows, or observes them in
different patterns.

The key difference from `anchored_mfa()` is that no single block is treated as
the privileged reference. The shared scores live in a latent reference row
space, and each block contributes through its own `row_index` mapping.

## What does a quick fit look like?

```{r load-packages}
library(muscal)
library(multivarious)
library(ggplot2)
```

```{r simulate-data, include = FALSE}
set.seed(42)
N <- 75
k <- 2
S_true <- qr.Q(qr(scale(matrix(rnorm(N * k), nrow = N, ncol = k), center = TRUE, scale = FALSE)))

idx1 <- sort(sample.int(N, 60, replace = FALSE))
idx2 <- sort(c(setdiff(seq_len(N), idx1), sample(idx1, 40, replace = FALSE)))
idx3 <- sort(sample.int(N, 55, replace = FALSE))

V1 <- svd(matrix(rnorm(20 * k), 20, k))$u[, seq_len(k), drop = FALSE]
V2 <- svd(matrix(rnorm(18 * k), 18, k))$u[, seq_len(k), drop = FALSE]
V3 <- svd(matrix(rnorm(16 * k), 16, k))$u[, seq_len(k), drop = FALSE]

X1 <- S_true[idx1, , drop = FALSE] %*% t(V1) +
  matrix(rnorm(length(idx1) * 20, sd = 0.1), nrow = length(idx1), ncol = 20)
X2 <- S_true[idx2, , drop = FALSE] %*% t(V2) +
  matrix(rnorm(length(idx2) * 18, sd = 0.1), nrow = length(idx2), ncol = 18)
X3 <- S_true[idx3, , drop = FALSE] %*% t(V3) +
  matrix(rnorm(length(idx3) * 16, sd = 0.1), nrow = length(idx3), ncol = 16)

blocks <- list(
  expression = scale(X1, center = TRUE, scale = FALSE),
  imaging = scale(X2, center = TRUE, scale = FALSE),
  behavior = scale(X3, center = TRUE, scale = FALSE)
)
row_index <- list(expression = idx1, imaging = idx2, behavior = idx3)
coverage <- tabulate(c(idx1, idx2, idx3), nbins = N)

full_blocks <- list(
  expression = scale(S_true %*% t(V1) + matrix(rnorm(N * 20, sd = 0.1), N, 20), center = TRUE, scale = FALSE),
  imaging = scale(S_true %*% t(V2) + matrix(rnorm(N * 18, sd = 0.1), N, 18), center = TRUE, scale = FALSE),
  behavior = scale(S_true %*% t(V3) + matrix(rnorm(N * 16, sd = 0.1), N, 16), center = TRUE, scale = FALSE)
)
```

```{r show-coverage}
c(
  n_expression_rows = nrow(blocks$expression),
  n_imaging_rows = nrow(blocks$imaging),
  n_behavior_rows = nrow(blocks$behavior)
)
```

```{r fit-aligned-mfa}
fit <- aligned_mfa(blocks, row_index, N = N, ncomp = 2, normalization = "None")
S <- multivarious::scores(fit)

stopifnot(all(dim(S) == c(N, 2)))
stopifnot(all(is.finite(S)))
stopifnot(length(fit$objective_trace) >= 1)
```

```{r latent-recovery-check}
cc <- cancor(
  scale(S_true, center = TRUE, scale = FALSE),
  scale(S, center = TRUE, scale = FALSE)
)$cor[1:2]

stopifnot(all(is.finite(cc)))
stopifnot(mean(cc) > 0.77)

round(cc, 3)
```

```{r quick-score-plot, echo = FALSE, fig.cap = "Aligned MFA estimates one score vector for each latent reference row. Here the points are colored by how many blocks cover that row."}
score_df <- data.frame(
  Comp1 = S[, 1],
  Comp2 = S[, 2],
  coverage = factor(coverage)
)

ggplot(score_df, aes(Comp1, Comp2, color = coverage)) +
  geom_point(alpha = 0.85, size = 2) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey75") +
  geom_vline(xintercept = 0, linetype = 2, color = "grey75") +
  labs(x = "Component 1", y = "Component 2", color = "Blocks")
```

Rows observed in two blocks are better constrained than rows observed once, but
the model still estimates a single coherent score space over all `N` latent
rows.

## How do row mappings work?

```{r inspect-row-index}
lapply(row_index, head, 8)
```

Each integer vector says which latent reference row each observed row belongs
to. This is the entire alignment mechanism: if row 7 in `expression` and row 3
in `imaging` both map to latent row 21, they contribute to the same shared
score vector.

## What happens when all rows line up?

When every block observes the same rows in the same order, `aligned_mfa()`
should agree with ordinary `mfa()`.

```{r reduction-to-mfa}
fit_aligned_full <- aligned_mfa(
  full_blocks,
  list(expression = seq_len(N), imaging = seq_len(N), behavior = seq_len(N)),
  N = N,
  ncomp = 2,
  normalization = "None",
  ridge = 1e-10,
  max_iter = 200,
  tol = 1e-10
)
if (requireNamespace("genpca", quietly = TRUE)) {
  fit_mfa <- mfa(full_blocks, ncomp = 2, normalization = "None")

  P1 <- multivarious::scores(fit_aligned_full) %*%
    solve(crossprod(multivarious::scores(fit_aligned_full)), t(multivarious::scores(fit_aligned_full)))
  P2 <- multivarious::scores(fit_mfa) %*%
    solve(crossprod(multivarious::scores(fit_mfa)), t(multivarious::scores(fit_mfa)))
  rel <- norm(P1 - P2, type = "F") / (norm(P2, type = "F") + 1e-12)

  stopifnot(is.finite(rel))
  stopifnot(rel < 0.05)

  rel
} else {
  "Package 'genpca' not installed; skipping direct MFA reduction check."
}
```

That reduction check is important: it confirms that you are getting a genuine
extension of MFA rather than a separate, incompatible factor model.

## What should you inspect while fitting?

```{r inspect-loadings}
sapply(fit$V_list, dim)
```

```{r objective-trace, echo = FALSE, fig.cap = "The aligned-MFA objective should fall quickly and then flatten."}
obj_df <- data.frame(
  iteration = seq_along(fit$objective_trace),
  objective = fit$objective_trace
)

ggplot(obj_df, aes(iteration, objective)) +
  geom_line(linewidth = 0.8, color = "#9c2f2f") +
  geom_point(size = 1.6, color = "#9c2f2f") +
  labs(x = "Iteration", y = "Objective")
```

The loading dimensions tell you how much information each block carries. The
objective trace tells you whether the alternating updates have settled cleanly.

## What about inference and validation?

`aligned_mfa()` already fits into the package's task-aware prediction surface:
it exposes aligned scores and reconstructions, so you can evaluate held-out
workflows with explicit `cv_muscal()` callbacks when you have a clear fold
construction strategy.

It also now exposes generic refit metadata, so `infer_muscal()` can be used for
bootstrap or permutation summaries. For this method, fit-quality statistics are
often more informative than raw component standard deviations, because the score
space is orthonormalized during fitting.

## Where next?

- `vignette("linked_mfa")` for the anchored version with an explicit reference block
- `vignette("mfa")` for the classical same-row setting
- `vignette("model_evaluation")` for the shared evaluation API across supported methods
- `?aligned_mfa` for `feature_groups`, `feature_lambda`, and weighting options
