---
title: Aligned Multiblock Canonical Correlation Analysis
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aligned Multiblock Canonical Correlation Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE) && exists("params")) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
```

## Why aligned MCCA?

`aligned_mcca()` solves the same row-mismatch problem as `aligned_mfa()`, but in
the language of canonical correlation. You use it when each block has its own
row set, yet all blocks should still line up on one shared score space over a
common latent population.

It is the correlation-based counterpart to `aligned_mfa()`: same alignment
idea, different objective.

## What does a quick fit look like?

```{r load-packages}
library(muscal)
library(multivarious)
library(ggplot2)
```

```{r simulate-data}
set.seed(42)
N <- 75
k <- 2
Z <- scale(matrix(rnorm(N * k), nrow = N, ncol = k), center = TRUE, scale = FALSE)

idx1 <- sort(sample.int(N, 60, replace = FALSE))
idx2 <- sort(c(setdiff(seq_len(N), idx1), sample(idx1, 40, replace = FALSE)))
idx3 <- sort(sample.int(N, 55, replace = FALSE))

make_block <- function(idx, p, noise_sd = 0.1) {
  A <- matrix(rnorm(p * k), nrow = p, ncol = k)
  X <- Z[idx, , drop = FALSE] %*% t(A) +
    matrix(rnorm(length(idx) * p, sd = noise_sd), nrow = length(idx), ncol = p)
  scale(X, center = TRUE, scale = FALSE)
}

blocks <- list(
  transcriptomics = make_block(idx1, 20),
  imaging = make_block(idx2, 22),
  physiology = make_block(idx3, 18)
)
row_index <- list(transcriptomics = idx1, imaging = idx2, physiology = idx3)
coverage <- tabulate(c(idx1, idx2, idx3), nbins = N)

full_blocks <- list(
  transcriptomics = make_block(seq_len(N), 20),
  imaging = make_block(seq_len(N), 22)
)
```

```{r show-coverage}
c(
  n_transcriptomics_rows = nrow(blocks$transcriptomics),
  n_imaging_rows = nrow(blocks$imaging),
  n_physiology_rows = nrow(blocks$physiology)
)
```

```{r fit-aligned-mcca}
fit <- aligned_mcca(blocks, row_index, N = N, ncomp = 2, ridge = 1e-6)
S <- multivarious::scores(fit)

stopifnot(all(dim(S) == c(N, 2)))
stopifnot(all(is.finite(S)))
```

```{r latent-recovery-check}
cc <- cancor(
  scale(Z, center = TRUE, scale = FALSE),
  scale(S, center = TRUE, scale = FALSE)
)$cor[1:2]

stopifnot(all(is.finite(cc)))
stopifnot(mean(cc) > 0.83)

round(cc, 3)
```

```{r quick-score-plot, echo = FALSE, fig.cap = "Aligned MCCA scores over the latent reference rows. Darker coverage means more blocks contribute to that row."}
score_df <- data.frame(
  Comp1 = S[, 1],
  Comp2 = S[, 2],
  coverage = factor(coverage)
)

ggplot(score_df, aes(Comp1, Comp2, color = coverage)) +
  geom_point(alpha = 0.85, size = 2) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey75") +
  geom_vline(xintercept = 0, linetype = 2, color = "grey75") +
  labs(x = "Component 1", y = "Component 2", color = "Blocks")
```

The shared components recover the latent signal even though no block covers the
same row pattern as any other block.

## How is this different from ordinary MCCA?

When all blocks share the same rows, `aligned_mcca()` should collapse back to
`mcca()`.

```{r reduction-to-mcca}
fit_aligned_full <- aligned_mcca(
  full_blocks,
  list(transcriptomics = seq_len(N), imaging = seq_len(N)),
  N = N,
  ncomp = 2,
  ridge = 1e-6
)
fit_mcca <- mcca(full_blocks, ncomp = 2, ridge = 1e-6)

P1 <- multivarious::scores(fit_aligned_full) %*%
  solve(crossprod(multivarious::scores(fit_aligned_full)), t(multivarious::scores(fit_aligned_full)))
P2 <- multivarious::scores(fit_mcca) %*%
  solve(crossprod(multivarious::scores(fit_mcca)), t(multivarious::scores(fit_mcca)))
rel <- norm(P1 - P2, type = "F") / (norm(P2, type = "F") + 1e-12)

stopifnot(is.finite(rel))
stopifnot(rel < 1e-8)

rel
```

That reduction matters because it tells you the alignment layer is only solving
the missing-row problem. It is not changing the core GCCA objective when the
problem is already fully aligned.

## What should you inspect after fitting?

```{r inspect-partial-scores}
sapply(fit$partial_scores, dim)
```

```{r inspect-canonical-weights}
sapply(fit$canonical_weights, dim)
```

The `partial_scores` live at the observed-row level for each block. The
`canonical_weights` live in each block's feature space. Together they tell you
how the shared reference-space scores are supported by the observed data.

## When should you use aligned MCCA?

Use `aligned_mcca()` when:

- each block has its own row set,
- you want correlation-based shared components,
- repeated or subset row mappings are part of the data collection process, and
- no block should be treated as the anchor.

If you *do* have a fully observed anchor block, `anchored_mcca()` is the more
direct interface.

## Where next?

- `?aligned_mcca` for weighting and ridge options
- `?anchored_mcca` for the anchored wrapper
- `vignette("aligned_mfa")` for the variance-based aligned alternative
