---
title: Bidirectional Response Alignment with aligned_interbattery()
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bidirectional Response Alignment with aligned_interbattery()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE) && exists("params")) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
```

## Why use `aligned_interbattery()`?

Use `aligned_interbattery()` when you have two paired block families for the
same subjects and neither side is the privileged response surface. You want a
shared common space, but you also want to predict from `X` to `Y`, from `Y` to
`X`, or complete a partially observed bundle using both sides together.

The key contract is symmetric: each subject gets common scores plus one score
system for the `X` side and one for the `Y` side. That makes this a better fit
than forcing one side to play the role of "predictors" and the other to play
the role of "responses."

## What does a quick fit look like?

```{r load-packages}
library(muscal)
library(knitr)
```

```{r simulate-data, include = FALSE}
.mean_sq_error <- function(pred, truth) {
  mean((pred - truth)^2)
}

.sim_aib_subject <- function(N, P1, P2, Q1, noise = 0.04, include_d2 = TRUE) {
  K <- ncol(P1)
  T <- qr.Q(qr(matrix(rnorm(N * K), N, K)))
  U <- T + matrix(rnorm(N * K, sd = noise), N, K)
  V <- T + matrix(rnorm(N * K, sd = noise), N, K)

  X <- list(
    D1 = U %*% t(P1) + matrix(rnorm(N * nrow(P1), sd = noise), N, nrow(P1))
  )
  if (isTRUE(include_d2)) {
    X$D2 <- U %*% t(P2) + matrix(rnorm(N * nrow(P2), sd = noise), N, nrow(P2))
  }

  Y <- list(
    E1 = V %*% t(Q1) + matrix(rnorm(N * nrow(Q1), sd = noise), N, nrow(Q1))
  )

  list(X = X, Y = Y)
}

.sim_aib_data <- function(subject_sizes = c(S1 = 14, S2 = 13, S3 = 12),
                          noise = 0.04,
                          seed = 11) {
  set.seed(seed)
  K <- 2L
  P1 <- matrix(rnorm(5 * K), 5, K)
  P2 <- matrix(rnorm(4 * K), 4, K)
  Q1 <- matrix(rnorm(3 * K), 3, K)

  X <- list()
  Y <- list()
  for (subj in names(subject_sizes)) {
    sim_subj <- .sim_aib_subject(
      N = as.integer(subject_sizes[[subj]]),
      P1 = P1,
      P2 = P2,
      Q1 = Q1,
      noise = noise
    )
    X[[subj]] <- sim_subj$X
    Y[[subj]] <- sim_subj$Y
  }

  list(
    X = X,
    Y = Y,
    params = list(P1 = P1, P2 = P2, Q1 = Q1, K = K)
  )
}

train <- .sim_aib_data()
new_subject <- .sim_aib_subject(
  N = 9,
  P1 = train$params$P1,
  P2 = train$params$P2,
  Q1 = train$params$Q1,
  noise = 0.04
)
partial_both <- list(
  x = list(D1 = new_subject$X$D1),
  y = list(E1 = new_subject$Y$E1)
)
```

```{r fit-aligned-interbattery}
fit <- aligned_interbattery(
  X = train$X,
  Y = train$Y,
  ncomp = 2,
  max_iter = 10
)
```

```{r quick-checks, include = FALSE}
pred_y <- predict(
  fit,
  new_subject$X,
  from = "x",
  type = "prediction",
  target_block = "E1"
)
pred_x <- predict(
  fit,
  new_subject$Y$E1,
  block = "E1",
  from = "y",
  type = "prediction",
  target_block = "D1"
)
pred_d2 <- predict(
  fit,
  partial_both,
  from = "both",
  type = "prediction",
  target_block = "x:D2"
)
proj_both <- project(
  fit,
  partial_both,
  from = "both",
  side = "both"
)

quick_summary <- data.frame(
  path = c(
    "x:D1 + x:D2 -> y:E1",
    "y:E1 -> x:D1",
    "x:D1 + y:E1 -> x:D2"
  ),
  mean_test_mse = round(c(
    .mean_sq_error(pred_y, new_subject$Y$E1),
    .mean_sq_error(pred_x, new_subject$X$D1),
    .mean_sq_error(pred_d2, new_subject$X$D2)
  ), 4)
)

stopifnot(all(is.finite(quick_summary$mean_test_mse)))
stopifnot(all(dim(proj_both$common) == c(9, 2)))
stopifnot(all(dim(proj_both$x) == c(9, 2)))
stopifnot(all(dim(proj_both$y) == c(9, 2)))
stopifnot(identical(fit$task, "bidirectional_prediction"))
```

```{r quick-summary}
kable(quick_summary, align = c("l", "r"))
```

The important point is not which direction wins on this toy simulation. It is
that one fitted object supports all three contracts: `X -> Y`, `Y -> X`, and
joint completion from partially observed bundles.

## What do the inputs look like?

You can supply nested `subject -> domain` lists on both sides. Blocks with the
same domain name and column schema share loadings across subjects, which is the
main way the model learns a common geometry rather than fitting each subject
independently.

```{r input-shape}
c(
  n_subjects = length(train$X),
  x_domains_per_subject = length(train$X$S1),
  y_domains_per_subject = length(train$Y$S1)
)
```

Here each subject has two `X`-side domains (`D1`, `D2`) and one `Y`-side domain
(`E1`). In a real study those domains might be assays, feature families, or
paired modalities collected on the same rows.

## How do you project from one side or both?

If you observe both sides for the same new rows, `project()` can return the
common scores together with the side-specific score systems.

```{r project-both}
common_scores <- project(
  fit,
  partial_both,
  from = "both",
  side = "common"
)

dim(common_scores)
```

If you only want to reconstruct the blocks you actually observed, use
`type = "reconstruction"`.

```{r reconstruct-observed}
reconstructed <- predict(
  fit,
  partial_both,
  from = "both",
  type = "reconstruction"
)

names(reconstructed)
```

That separation matters. Reconstruction is for the observed side bundle.
Prediction is for completing the opposite side or an unobserved block given the
common space you inferred from what you have.

## Where next?

- `vignette("response_alignment")` when one side really is the privileged
  response surface
- `?aligned_interbattery` for the fitting contract and supported input layouts
- `?project.aligned_interbattery` and `?predict.aligned_interbattery` for the
  out-of-sample projection and completion rules
