---
title: Barycentric MFA (BaMFA)
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Barycentric MFA (BaMFA)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE) && exists("params")) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
```

## Why BaMFA?

Multi-subject data often contains two kinds of structure: patterns shared
across all subjects (*global* components) and patterns unique to individuals
(*local* components). Standard MFA captures the global structure but discards
subject-specific variation. If you care about individual differences — which
subjects deviate, and how — you need a method that explicitly models both.

BaMFA (Barycentric Multiple Factor Analysis) decomposes multi-block data into
**global** and **local** components simultaneously. Global components capture
the consensus structure; local components capture what makes each subject
unique.

## Quick start

```{r load-packages}
library(muscal)
library(multivarious)
library(ggplot2)
```

We simulate 6 subjects with shared structure (3 global factors) plus
subject-specific structure (2 local factors per subject).

```{r simulate-data, include = FALSE}
set.seed(42)
n_subjects <- 6
n_obs <- 50
n_vars <- 20
r_global <- 3
r_local <- 2

# Shared global loading structure
V_global <- svd(matrix(rnorm(n_vars * r_global), n_vars, r_global))$u

# Shared global scores
F_global <- qr.Q(qr(scale(matrix(rnorm(n_obs * r_global), n_obs, r_global),
                           scale = FALSE)))

data_list <- lapply(seq_len(n_subjects), function(i) {
  # Subject-specific local loadings (orthogonal to global)
  V_local_raw <- matrix(rnorm(n_vars * r_local), n_vars, r_local)
  # Project out global subspace to make local orthogonal-ish
  V_local_raw <- V_local_raw - V_global %*% (t(V_global) %*% V_local_raw)
  V_local <- svd(V_local_raw)$u[, seq_len(r_local)]

  F_local <- matrix(rnorm(n_obs * r_local, sd = 0.5), n_obs, r_local)

  X <- F_global %*% t(V_global) + F_local %*% t(V_local) +
    matrix(rnorm(n_obs * n_vars, sd = 0.3), n_obs, n_vars)
  scale(X, scale = FALSE)
})
names(data_list) <- paste0("Subject_", seq_len(n_subjects))
```

```{r show-data}
length(data_list)
sapply(data_list, dim)
```

Fit BaMFA with 3 global and 2 local components:

```{r fit-bamfa}
fit <- bamfa(data_list, k_g = 3, k_l = 2, niter = 20)
fit
```

```{r quick-score-plot, echo = FALSE, fig.cap = "Mean global scores across subjects. These capture the consensus structure shared by all blocks."}
autoplot(fit, block = "mean")
```

## Global vs. local components

The key idea: BaMFA decomposes each block's data as:

$$\mathbf{X}_i \approx \mathbf{S}_i \mathbf{B}_i^\top + \mathbf{U}_i \mathbf{V}_i^\top$$

where:

- $\mathbf{S}_i$ are the **global scores** — similar across subjects
- $\mathbf{B}_i$ are the **global loadings**
- $\mathbf{U}_i$ are the **local scores** — unique to subject $i$
- $\mathbf{V}_i$ are the **local loadings**

The global loadings $\mathbf{B}_i$ are encouraged to be similar across
subjects, while the local loadings $\mathbf{V}_i$ are free to differ.

## Examining the components

```{r global-components}
# Global loadings: one per block, should be similar
sapply(fit$B_list, dim)

# Local loadings: one per block, capture individual structure
sapply(fit$V_list, dim)
```

## Convergence

```{r convergence, fig.cap = "BaMFA convergence trace. The alternating least squares algorithm typically converges within 10-20 iterations."}
plot_convergence(fit)
```

## Per-subject score plots

View the global scores for a specific subject:

```{r subject-scores, fig.cap = "Global scores for Subject 1."}
autoplot(fit, block = 1)
```

## Partial factor scores

Compare how the global structure looks across subjects:

```{r partial-scores, fig.cap = "Partial factor scores: each subject's global scores overlaid. Agreement indicates consistent global structure.", fig.width = 7, fig.height = 5}
plot_partial_scores(fit, connect = TRUE, show_consensus = TRUE)
```

Short connecting lines mean subjects agree about the global structure of
an observation. Large deviations indicate observations where subjects diverge.

## The local penalty (lambda_l)

The `lambda_l` parameter controls regularization on local components. By
default it is 0 (no regularization). Increasing it shrinks local components,
pushing more structure into the global subspace:

```{r lambda-l, eval = FALSE}
# Strong local regularization — most structure forced into global
fit_sparse <- bamfa(data_list, k_g = 3, k_l = 2, niter = 20, lambda_l = 1)
```

## Choosing k_g and k_l

- **`k_g`** (global components): Set this to the number of shared factors you
  expect. Start with the number of dominant eigenvalues from pooled PCA.
- **`k_l`** (local components): Set this to capture meaningful individual
  variation. Too many local components can absorb noise.

A useful diagnostic: if partial factor scores are tight (short lines in the
plot), `k_g` is capturing the shared structure well. If they're spread out,
you may need more global components.

## When to use BaMFA

BaMFA is appropriate when:

- You have multi-subject or multi-condition data blocks
- You want to separate shared (global) from individual (local) structure
- Individual differences are scientifically interesting, not just noise
- You need interpretable per-subject components

BaMFA is *not* appropriate when:

- You only care about the consensus (use `mfa()` — simpler and faster)
- Blocks have different observations (use `anchored_mfa()`)
- Your data are covariance matrices (use `covstatis()`)

## Next steps

- `vignette("mfa")` — Standard MFA for consensus-only analysis
- `vignette("penalized_mfa")` — Penalized MFA for loading similarity
- `?bamfa` — Full parameter documentation
