---
title: COVSTATIS Analysis
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{COVSTATIS Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE) && exists("params")) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
```

## Why COVSTATIS?

You have correlation or covariance matrices from multiple subjects — say,
resting-state functional connectivity measured in 10 people — and you want to
find the shared structure. Which brain regions co-activate consistently across
subjects? Which subjects are outliers? COVSTATIS answers these questions by
computing a *compromise* matrix that optimally represents the set, then
projecting both variables (ROIs) and subjects into a common low-dimensional
space.

COVSTATIS is the covariance-matrix variant of STATIS, a "French school"
method for analyzing collections of data tables.

## Quick start

```{r load-packages}
library(muscal)
library(multivarious)
library(ggplot2)
```

We simulate 8 subjects, each with a 12 x 12 correlation matrix. All subjects
share the same underlying factor structure, but with individual noise.

```{r simulate-data, include = FALSE}
set.seed(42)
n_subjects <- 8
n_vars <- 12
n_obs <- 100

# Shared loading structure (3 latent factors)
L <- matrix(rnorm(n_vars * 3), n_vars, 3)
shared_cov <- L %*% t(L) + 0.5 * diag(n_vars)
R <- chol(shared_cov)

# Per-subject correlation matrices
cor_list <- lapply(seq_len(n_subjects), function(i) {
  X <- matrix(rnorm(n_obs * n_vars), n_obs, n_vars) %*% R
  # Add subject-specific signal shift
  X <- X + matrix(rnorm(n_obs * n_vars, sd = 0.2 * i), n_obs, n_vars)
  cor(X)
})
names(cor_list) <- paste0("Subj_", seq_len(n_subjects))
```

```{r show-data}
# 8 subjects, each with a 12 x 12 correlation matrix
length(cor_list)
dim(cor_list[[1]])
```

Fit COVSTATIS and extract the compromise:

```{r fit-covstatis}
fit <- covstatis(cor_list, ncomp = 3)
```

```{r roi-scores, echo = FALSE, fig.cap = "ROI scores in the compromise space. Variables that co-vary consistently across subjects cluster together."}
autoplot(fit, type = "roi")
```

The ROI score plot shows `r nrow(fit$roi_scores)` variables positioned
according to their role in the compromise. Variables close together have
similar correlation patterns across subjects.

## How COVSTATIS works

The algorithm has four steps:

1. **Preprocess** each matrix — optionally double-center (Gower transformation)
   and normalize to unit Frobenius norm.

2. **Compute the RV matrix** — a matrix of pairwise similarities between
   the input matrices (analogous to correlations between subjects).

3. **Determine optimal weights** (`alpha`) from the first eigenvector of the
   RV matrix. Subjects whose matrices are more "typical" receive higher weight.

4. **Form the compromise** as a weighted sum of the normalized matrices,
   then eigendecompose it to get scores and loadings.

## Subject-level analysis

Each subject contributes differently to the compromise. The `alpha` weights
reveal who is most representative:

```{r subject-weights}
fit$alpha
```

```{r block-weights-plot, echo = FALSE, fig.cap = "Block weights (alpha). Higher weight means the subject's matrix is more representative of the group."}
plot_block_weights(fit)
```

You can also view subjects in the compromise space:

```{r subject-scores, fig.cap = "Subject scores. Subjects close together have similar correlation structure."}
autoplot(fit, type = "subjects")
```

## Variance explained

```{r variance, fig.cap = "Variance explained by each compromise component."}
plot_variance(fit, type = "bar")
```

## Partial factor scores

Partial scores show how each subject "sees" the ROIs. Lines connect each
subject's view to the consensus:

```{r partial-scores, fig.cap = "Partial factor scores. Each color is one subject; open circles show the consensus. Short connecting lines indicate agreement.", fig.width = 7, fig.height = 5}
plot_partial_scores(fit, connect = TRUE, show_consensus = TRUE)
```

## Block similarity

The RV matrix shows pairwise similarity between subjects:

```{r block-similarity, fig.cap = "RV coefficient matrix between subjects. Values near 1 mean similar correlation structure."}
plot_block_similarity(fit)
```

## Projecting new subjects

After fitting, you can project a new subject's correlation matrix into the
compromise space without refitting:

```{r project-new, include = FALSE}
# Generate a new subject
X_new <- matrix(rnorm(n_obs * n_vars), n_obs, n_vars) %*% R
X_new <- X_new + matrix(rnorm(n_obs * n_vars, sd = 0.15), n_obs, n_vars)
new_cor <- cor(X_new)
```

```{r project-cov}
proj <- project_cov(fit, new_cor)
dim(proj)
head(proj, 4)
```

For a richer analysis of new subjects — including RV coefficients and distance
to the compromise — use `project_subjects()`:

```{r project-subjects}
result <- project_subjects(fit, list(new_cor), subject_ids = "New_Subject")
result$scalar_summaries
```

## Projecting a covariate

You can ask: "which compromise dimensions relate to an external variable?"
For example, projecting a subject-level cognitive score:

```{r project-covariate}
# Simulated covariate (e.g., memory score per subject)
y <- rnorm(n_subjects)
dim_cos <- project_covariate(fit, y, what = "dimension", scale = "cosine")
dim_cos
```

Values near +/- 1 indicate that a compromise dimension strongly tracks the
covariate.

## When to use COVSTATIS

COVSTATIS is appropriate when:

- Your data are square symmetric matrices (correlations, covariances,
  distance matrices)
- You want to find a consensus structure across multiple tables
- You need to identify outlier subjects or conditions
- You want to project new observations or external covariates

COVSTATIS is *not* appropriate when:

- Your data are raw observation-by-variable matrices (use `mfa()` instead)
- You need supervised discrimination (use `bada()` or `bamfa()`)

## Next steps

- `vignette("mfa")` — MFA for raw data blocks
- `?project_subjects` — Detailed new-subject projection
- `?project_covariate` — Supplementary variable analysis

## References

Abdi, H., Williams, L. J., Valentin, D., & Bennani-Dosse, M. (2012).
STATIS and DISTATIS: optimum multitable principal component analysis and
three way metric multidimensional scaling. *WIREs Computational Statistics*,
4(2), 124-167.
