---
title: Integrative PCA (iPCA)
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrative PCA (iPCA)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE) && exists("params")) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
```

## Why iPCA?

When you run PCA on concatenated multi-block data, blocks with more variables
or higher variance dominate the result. MFA addresses this by normalizing each
block's first singular value to 1. But this is a single global correction — it
doesn't adapt to the actual signal content of each block.

Integrative PCA (iPCA) takes a different approach. It applies per-block
*multiplicative* penalties that continuously reweight each block's contribution
based on how well it aligns with the emerging shared structure. The penalty
strength `lambda` controls how aggressively blocks are equalized: small
`lambda` approaches standard PCA, large `lambda` gives each block equal
influence regardless of size.

iPCA was introduced by Tang and Allen (2021) and implemented here using an
efficient Flip-Flop algorithm.

## Quick start

```{r load-packages}
library(muscal)
library(multivarious)
library(ggplot2)
```

```{r simulate-data}
sim <- synthetic_multiblock(
  S = 4, n = 60,
  p = c(20, 30, 15, 25),
  r = 3, sigma = 0.3, seed = 42
)
sapply(sim$data_list, dim)
```

Fit iPCA with a moderate penalty:

```{r fit-ipca}
fit <- ipca(sim$data_list, ncomp = 3, lambda = 1)
S <- scores(fit)
dim(S)
```

```{r quick-score-plot, echo = FALSE, fig.cap = "iPCA scores. The multiplicative penalty balances block contributions while preserving shared structure."}
df <- data.frame(PC1 = S[, 1], PC2 = S[, 2])
ggplot(df, aes(PC1, PC2)) +
  geom_point(color = "steelblue", alpha = 0.8, size = 2) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey70") +
  geom_vline(xintercept = 0, linetype = 2, color = "grey70") +
  labs(title = "iPCA Score Plot",
       x = sprintf("Component 1 (sd = %.2f)", fit$sdev[1]),
       y = sprintf("Component 2 (sd = %.2f)", fit$sdev[2]))
```

## How iPCA works

iPCA solves a penalized eigenvalue problem. For $K$ blocks
$\mathbf{X}_1, \ldots, \mathbf{X}_K$, it finds shared eigenvectors $u$ by
maximizing:

$$\sum_{k=1}^K f_\lambda\!\left(\frac{u^\top \mathbf{X}_k^\top \mathbf{X}_k\, u}
{\mathrm{tr}(\mathbf{X}_k^\top \mathbf{X}_k)}\right)$$

where $f_\lambda$ is a concave penalty function (the multiplicative Frobenius
penalty). This upweights blocks that align well with the current direction and
downweights those that don't — a soft, adaptive form of integration.

## The lambda parameter

`lambda` controls the integration strength:

- **`lambda` near 0**: Standard PCA on concatenated data (no reweighting).
  Blocks with more variables dominate.
- **`lambda = 1`**: Moderate integration. Blocks are softly equalized.
- **Large `lambda`** (10+): Strong equalization. Every block contributes
  roughly equally, regardless of size.

```{r lambda-comparison}
fit_low  <- ipca(sim$data_list, ncomp = 3, lambda = 0.01)
fit_high <- ipca(sim$data_list, ncomp = 3, lambda = 10)
```

```{r lambda-plot, echo = FALSE, fig.cap = "Effect of lambda on the score space. Low lambda (left) lets larger blocks dominate; high lambda (right) equalizes contributions.", fig.width = 8, fig.height = 4}
S_lo <- scores(fit_low)
S_hi <- scores(fit_high)
df_lambda <- rbind(
  data.frame(PC1 = S_lo[, 1], PC2 = S_lo[, 2], lambda = "lambda == 0.01"),
  data.frame(PC1 = S_hi[, 1], PC2 = S_hi[, 2], lambda = "lambda == 10")
)
df_lambda$lambda <- factor(df_lambda$lambda, levels = c("lambda == 0.01", "lambda == 10"))
ggplot(df_lambda, aes(PC1, PC2)) +
  geom_point(color = "steelblue", alpha = 0.8, size = 2) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey70") +
  geom_vline(xintercept = 0, linetype = 2, color = "grey70") +
  facet_wrap(~ lambda, labeller = label_parsed) +
  labs(x = "Component 1", y = "Component 2")
```

## Tuning lambda with held-out data

`ipca_tune_alpha()` selects the optimal penalty by holding out a fraction of
the data and measuring reconstruction error:

```{r tune-alpha}
tune <- ipca_tune_alpha(
  sim$data_list, ncomp = 3,
  alpha_grid = c(0.01, 0.1, 1, 10),
  holdout_frac = 0.1
)
tune$best_alpha
```

```{r tune-plot, echo = FALSE, fig.cap = "Held-out MSE across alpha (lambda) values. Lower is better."}
df_tune <- data.frame(alpha = tune$results$alpha, mse = tune$results$holdout_mse)
ggplot(df_tune, aes(alpha, mse)) +
  geom_line(color = "steelblue") +
  geom_point(color = "steelblue", size = 3) +
  geom_vline(xintercept = tune$best_alpha, linetype = 2, color = "firebrick") +
  scale_x_log10() +
  labs(x = expression(alpha ~ (lambda)), y = "Held-out MSE",
       title = "iPCA Tuning Curve")
```

The tuning result includes a refit on the full data at the best alpha:

```{r tune-refit}
fit_tuned <- tune$fit
dim(scores(fit_tuned))
```

## Partial scores

Each block contributes its own view of the observations through per-block
loadings:

```{r partial-scores}
# Per-block scores
ps <- fit$partial_scores
sapply(ps, dim)
```

## Projection

Project new observations onto the fitted model. You pass a list of blocks
with the same structure as the training data:

```{r project-new}
new_blocks <- lapply(sim$data_list, function(X) X[1:5, , drop = FALSE])
proj <- project(fit, new_blocks)
dim(proj)
```

## Inference and validation

iPCA now participates in the same generic evaluation surface as MFA. Use
`infer_muscal()` for resampling-based summaries, and `cv_muscal()` when you
want explicit fold-wise reconstruction metrics.

```{r ipca-bootstrap}
boot <- infer_muscal(
  fit,
  method = "bootstrap",
  statistic = "sdev",
  nrep = 6,
  seed = 303
)

boot$summary
```

```{r ipca-permutation}
perm <- infer_muscal(
  fit,
  method = "permutation",
  statistic = "sdev",
  nrep = 9,
  seed = 404
)

perm$component_results
```

```{r ipca-inference-checks}
stopifnot(all(is.finite(boot$summary$mean)))
stopifnot(all(boot$summary$upper >= boot$summary$lower))
stopifnot(all(perm$component_results$p_value >= 0))
stopifnot(all(perm$component_results$p_value <= 1))
```

For a package-level walkthrough of `infer_muscal()`, `cv_muscal()`, and the
task-aware metric registry, see `vignette("model_evaluation")`.

## Computation methods

iPCA supports two internal methods:

- **`"gram"`** (default for wide data): Works in sample space ($n \times n$).
  Efficient when blocks have more variables than observations.
- **`"dense"`**: Works in variable space. Better for tall data.

The `"auto"` setting (default) picks the right method based on data dimensions.

```{r method, eval = FALSE}
# Force gram mode for high-dimensional blocks
fit_gram <- ipca(sim$data_list, ncomp = 3, lambda = 1, method = "gram")
```

## When to use iPCA

iPCA is appropriate when:

- You have multiple data blocks on the same observations
- You want adaptive, data-driven integration (vs. MFA's fixed normalization)
- Blocks vary substantially in dimensionality or signal strength
- You want a tunable integration strength

iPCA is *not* appropriate when:

- Blocks have different observations (use `anchored_mfa()`)
- You need classical MFA block weights for interpretation (use `mfa()`)
- Your data are covariance matrices (use `covstatis()`)

## Next steps

- `vignette("mfa")` — Classical MFA with fixed normalization
- `vignette("model_evaluation")` — Generic inference and held-out evaluation
  workflows
- `vignette("penalized_mfa")` — MFA with loading similarity penalties
- `?ipca_tune_alpha` — Detailed tuning options

## References

Tang, T. M., & Allen, G. I. (2021). Integrated principal components analysis.
*Journal of Machine Learning Research*, 22(224), 1-71.
