---
title: Anchored Multiple Factor Analysis
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Anchored Multiple Factor Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE) && exists("params")) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
```

## Why Anchored MFA?

Standard MFA requires all blocks to share the same observations. This is often
unrealistic. Consider neuroimaging: you have a reference dataset (Y) with
structural measures on N subjects, but your auxiliary data — task fMRI,
resting-state, behavioral assessments — are available for overlapping but
different subsets of those subjects. Some subjects completed one scan, others
completed two, and the overlap varies by modality.

Anchored MFA solves this. It learns a shared score matrix for the reference
observations while allowing each auxiliary block to have its own row structure.
You simply specify which rows of each auxiliary block correspond to which rows
of the reference.

> **Naming note:** `anchored_mfa()` is the primary function. `linked_mfa()` is
> retained as a convenience alias.

## Quick start

```{r load-packages}
library(muscal)
library(multivarious)
library(ggplot2)
```

We create a reference block with 80 subjects and two auxiliary blocks measured
on overlapping subsets. All three share the same underlying latent structure so
the model has real signal to recover.

```{r simulate-data, include = FALSE}
set.seed(42)
N <- 80
r <- 3

# Shared latent scores
F_true <- matrix(rnorm(N * r), N, r)
F_true <- qr.Q(qr(scale(F_true, scale = FALSE)))

# Reference block: all N subjects, 20 variables
B_ref <- svd(matrix(rnorm(20 * r), 20, r))$u
Y <- F_true %*% t(B_ref) + 0.2 * matrix(rnorm(N * 20), N, 20)
Y <- scale(Y, scale = FALSE)

# Auxiliary block 1: 50 subjects (subset), 30 variables
idx1 <- sort(sample.int(N, 50))
V1 <- svd(matrix(rnorm(30 * r), 30, r))$u
X1 <- F_true[idx1, ] %*% t(V1) + 0.2 * matrix(rnorm(50 * 30), 50, 30)
X1 <- scale(X1, scale = FALSE)

# Auxiliary block 2: 40 subjects (different subset), 25 variables
idx2 <- sort(sample.int(N, 40))
V2 <- svd(matrix(rnorm(25 * r), 25, r))$u
X2 <- F_true[idx2, ] %*% t(V2) + 0.2 * matrix(rnorm(40 * 25), 40, 25)
X2 <- scale(X2, scale = FALSE)
```

```{r show-data-dims}
# Reference block: all 80 subjects
dim(Y)

# Auxiliary blocks: overlapping subsets
dim(X1)  # 50 subjects
dim(X2)  # 40 subjects

# How many subjects overlap between auxiliary blocks?
length(intersect(idx1, idx2))
```

```{r fit-anchored}
fit <- anchored_mfa(
  Y = Y,
  X = list(X1 = X1, X2 = X2),
  row_index = list(X1 = idx1, X2 = idx2),
  ncomp = 3
)

# Scores are defined for all N reference subjects
S <- scores(fit)
dim(S)
```

```{r quick-score-plot, echo = FALSE, fig.cap = "Observations in the Anchored MFA shared score space, colored by block coverage. Darker points have more data sources informing their position."}
autoplot(fit, color = "coverage")
```

The score plot colors observations by how many blocks contribute to their
position. Subjects measured across more modalities have more information
constraining their scores.

## The model

Anchored MFA fits the structure:

$$Y \approx S B^\top$$
$$X_k \approx S[\mathrm{idx}_k,] V_k^\top$$

where:

- **Y** is the reference block (N x q)
- **S** is the shared score matrix (N x K)
- **B** contains the loadings for Y
- Each **X_k** is an auxiliary block with its own dimensions
- **V_k** contains the loadings for block k
- **idx_k** maps rows of X_k to rows of Y

The key insight: scores S live in the reference space (N rows), but each
auxiliary block contributes information through its row mapping.

## Row index mapping

The `row_index` argument is the core of Anchored MFA. Each element maps rows
of an auxiliary block to rows of the reference Y:

```{r mapping-example, eval = FALSE}
# row_index[[k]][i] says:
#   "row i of X[[k]] corresponds to row row_index[[k]][i] of Y"

# Subset data: 60 of 100 subjects have this modality
idx <- sample.int(100, 60, replace = FALSE)

# Repeated measures: subject 1 measured 3 times, subject 2 twice
idx_repeated <- c(rep(1L, 3), rep(2L, 2), 3L)
```

This flexibility handles subset data, repeated measures, and unbalanced
designs.

## Working with the results

The returned object behaves like a standard `multiblock_biprojector`:

```{r working-results}
# Global scores (N x ncomp)
dim(scores(fit))

# Block-specific loadings
dim(fit$B)              # Y loadings
sapply(fit$V_list, dim) # X_k loadings

# Block indices in concatenated space
str(fit$block_indices)
```

## Convergence diagnostics

Anchored MFA uses alternating least squares (ALS). Monitor convergence to
ensure the algorithm has settled:

```{r convergence, fig.cap = "Convergence trace of the ALS objective. A healthy trace decreases monotonically and flattens near the solution."}
plot_convergence(fit)
```

## Visualization

### Block coverage

The coverage heatmap shows which observations are present in which blocks —
essential for understanding your data's overlap structure:

```{r coverage, fig.cap = "Coverage heatmap: rows are observations, columns are blocks. Filled cells indicate presence.", fig.height = 5}
plot_coverage(fit)
```

### Score plot with coverage

Color observations by how many blocks contribute to their scores:

```{r autoplot-coverage, fig.cap = "Score plot colored by block coverage. Observations with more contributing blocks have richer information."}
autoplot(fit, color = "coverage")
```

### Partial factor scores

Compare how each block "sees" the observations:

```{r partial-scores, fig.cap = "Partial factor scores. Open circles mark the consensus position in S; colored points show each block's view.", fig.width = 7, fig.height = 5}
plot_partial_scores(fit, show_consensus = TRUE)
```

### Block weights and fit

```{r block-weights, fig.cap = "Block normalization weights."}
plot_block_weights(fit)
```

```{r block-fit, fig.cap = "Relative reconstruction quality per block. Higher R-squared means the model captures more of that block's variance."}
plot_block_fit(fit)
```

### Variable loadings

```{r loadings, fig.cap = "Top 15 variable loadings on component 1, colored by block."}
plot_loadings(fit, type = "bar", component = 1, top_n = 15)
```

## Block normalization

Like standard MFA, Anchored MFA supports block weighting:

- **`"MFA"`** (default): Inverse squared first singular value
- **`"None"`**: Uniform weights
- **`"custom"`**: User-supplied weights via `alpha`

```{r normalization-options, eval = FALSE}
# Custom weights: emphasize the reference block
fit_custom <- anchored_mfa(
  Y = Y,
  X = list(X1 = X1, X2 = X2),
  row_index = list(X1 = idx1, X2 = idx2),
  ncomp = 3,
  normalization = "custom",
  alpha = c(2, 1, 1)  # Y weight, X1 weight, X2 weight
)
```

## Feature similarity prior

When auxiliary blocks measure conceptually similar features — say, the same
brain regions across different tasks — you may want corresponding features to
have similar loadings. Anchored MFA supports this via a feature grouping prior.

```{r feature-groups, eval = FALSE}
# Automatic grouping: features with the same column name are grouped
fit_grouped <- anchored_mfa(
  Y = Y,
  X = list(X1 = X1, X2 = X2),
  row_index = list(X1 = idx1, X2 = idx2),
  ncomp = 3,
  feature_groups = "colnames",
  feature_lambda = 0.1
)

# Manual grouping via data frame
groups_df <- data.frame(
  block = c("X1", "X1", "X2", "X2"),
  feature = c(1, 5, 1, 3),
  group = c("ROI_A", "ROI_B", "ROI_A", "ROI_B"),
  weight = c(1, 1, 1, 1)
)

fit_grouped <- anchored_mfa(
  Y = Y,
  X = list(X1 = X1, X2 = X2),
  row_index = list(X1 = idx1, X2 = idx2),
  ncomp = 3,
  feature_groups = groups_df,
  feature_lambda = 0.5
)
```

The `feature_lambda` parameter controls the strength of the prior. Higher
values pull grouped features toward their shared center more strongly. Use
`plot_feature_groups()` to visualize the alignment.

## When to use Anchored MFA

Anchored MFA is appropriate when:

- You have a reference dataset with complete observations
- Auxiliary datasets are measured on overlapping but incomplete subsets
- You want to learn a shared representation in the reference space
- Features across blocks may have conceptual correspondence

Anchored MFA is *not* appropriate when:

- All blocks share identical observations (use `mfa()` instead)
- There is no natural reference block
- Block relationships are better captured by other methods (e.g., CCA for two
  blocks with different row and column spaces)

## Next steps

- `vignette("mfa")` — Standard MFA when all blocks share observations
- `?covstatis` — STATIS analysis for covariance matrices
- `?penalized_mfa` — MFA with sparsity penalties

## References

Anchored MFA extends the classical MFA framework to handle heterogeneous row
structures. The algorithm is based on weighted alternating least squares with
an optional quadratic group-shrinkage penalty on loadings.

Escofier, B., & Pagès, J. (1994). Multiple factor analysis (AFMULT package).
*Computational Statistics & Data Analysis*, 18(1), 121-140.
