---
title: Multiblock Canonical Correlation Analysis
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiblock Canonical Correlation Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE) && exists("params")) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
```

## Why MCCA?

Use `mcca()` when you have several blocks measured on the same observations and
you care about the *shared score space* rather than raw variance in the
concatenated data. The method looks for components that each block can project
onto coherently, so it is a natural choice when blocks differ in scale or
dimensionality but should still agree on the same latent structure.

If your main question is "what common signal do these blocks line up on?",
MCCA is often the right first model.

## What does a quick fit look like?

```{r load-packages}
library(muscal)
library(multivarious)
library(ggplot2)
```

```{r simulate-data, include = FALSE}
set.seed(42)
n <- 80
k <- 2
Z <- scale(matrix(rnorm(n * k), nrow = n, ncol = k), center = TRUE, scale = FALSE)

make_block <- function(p, noise_sd = 0.25) {
  A <- matrix(rnorm(p * k), nrow = p, ncol = k)
  X <- Z %*% t(A) + matrix(rnorm(n * p, sd = noise_sd), nrow = n, ncol = p)
  scale(X, center = TRUE, scale = FALSE)
}

blocks <- list(
  clinical = make_block(16),
  behavior = make_block(20),
  imaging = make_block(18)
)
latent_group <- factor(ifelse(Z[, 1] > 0, "positive latent score", "negative latent score"))
```

```{r show-blocks}
sapply(blocks, dim)
```

```{r fit-mcca}
fit <- mcca(blocks, ncomp = 2, ridge = 1e-6)
S <- multivarious::scores(fit)

stopifnot(all(dim(S) == c(n, 2)))
stopifnot(all(is.finite(S)))
```

```{r shared-signal-check}
cc <- cancor(
  scale(Z, center = TRUE, scale = FALSE),
  scale(S, center = TRUE, scale = FALSE)
)$cor[1:2]

stopifnot(all(is.finite(cc)))
stopifnot(min(cc) > 0.9)

round(cc, 3)
```

```{r quick-score-plot, echo = FALSE, fig.cap = "MCCA scores recover the shared low-dimensional structure. The coloring is derived from the first latent factor used to generate the blocks."}
score_df <- data.frame(
  Comp1 = S[, 1],
  Comp2 = S[, 2],
  latent_group = latent_group
)

ggplot(score_df, aes(Comp1, Comp2, color = latent_group)) +
  geom_point(alpha = 0.8, size = 2) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey75") +
  geom_vline(xintercept = 0, linetype = 2, color = "grey75") +
  labs(x = "Component 1", y = "Component 2", color = NULL)
```

The canonical correlations above are the first sanity check. In this simulated
example, the fitted scores recover the two shared latent directions very
strongly.

## What is the method actually doing?

MCCA builds a compromise score space that all blocks can project into. In the
MAXVAR formulation used here, each block contributes a ridge-stabilized
projection operator, and the shared components are the leading eigenvectors of
their weighted sum.

That is why `mcca()` stays stable even when some blocks are wide (`p >> n`) or
close to rank-deficient after centering.

## What should you inspect after fitting?

```{r inspect-partial-scores}
sapply(fit$partial_scores, dim)
```

```{r partial-score-check}
S_sum <- Reduce(`+`, fit$partial_scores)
max_diff <- max(abs(S_sum - S))

stopifnot(is.finite(max_diff))
stopifnot(max_diff < 1e-6)

max_diff
```

Each block has its own partial score matrix. Their sum reproduces the
compromise scores, so large disagreements across partial scores tell you where
the blocks are pulling in different directions.

## When should you reach for MCCA?

Use `mcca()` when:

- your blocks share the same rows,
- you want a common score space driven by cross-block agreement,
- some blocks are much wider than others, and
- you need ridge stabilization in high dimensions.

Use `mfa()` instead when you want a classical variance-balancing factor model
rather than a canonical-correlation view.

## Where next?

- `vignette("mfa")` for variance-based multiblock integration
- `vignette("aligned_mcca")` for the same idea when blocks do not share the same rows
- `?mcca` for tuning `ridge` and `block_weights`
