---
title: Model Evaluation and Inference
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Evaluation and Inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE) && exists("params")) ggplot2::theme_set(albersdown::theme_albers(params$family))
CAN_EVAL <- requireNamespace("genpca", quietly = TRUE) &&
  requireNamespace("multidesign", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  eval = CAN_EVAL
)
```

## Why evaluate a multiblock model?

Exploratory score plots are useful, but they are not the whole story. In
`muscal`, the main evaluation questions are usually:

- Are the leading components stable under resampling?
- Is the shared structure stronger than a row-shuffled null?
- How well does the fitted model reconstruct or predict held-out data?

This vignette shows the package-level tools for answering those questions. The
same workflow applies across methods that expose a standard fit contract.

## Load the package

```{r load-packages}
library(muscal)
library(multivarious)
library(ggplot2)
```

## What does the evaluation API cover?

`muscal` exposes one generic resampling helper and one generic CV helper:

```{r metric-overview}
metric_registry()
default_metrics("reconstruction")
default_metrics("response_prediction")
```

The registered task families are:

- `reconstruction` for same-row multiblock models such as `mfa()`
- `response_prediction` for anchored models that predict a reference block
- `retrieval_alignment` for row-alignment style retrieval tasks

## Bootstrap and permutation inference for MFA

We start with a standard same-row MFA problem.

```{r simulate-mfa}
sim <- synthetic_multiblock(
  S = 3, n = 48,
  p = c(12, 10, 8),
  r = 2, sigma = 0.25, seed = 7
)

fit_mfa <- mfa(sim$data_list, ncomp = 2)

stopifnot(all(is.finite(scores(fit_mfa))))
stopifnot(length(fit_mfa$sdev) == 2)
```

### Bootstrap component stability

Bootstrap inference is useful when you want interval estimates for the leading
component strengths.

```{r mfa-bootstrap}
boot_mfa <- infer_muscal(
  fit_mfa,
  method = "bootstrap",
  statistic = "sdev",
  nrep = 8,
  seed = 101
)

boot_mfa$summary
```

```{r mfa-bootstrap-check}
stopifnot(all(is.finite(boot_mfa$summary$mean)))
stopifnot(all(boot_mfa$summary$upper >= boot_mfa$summary$lower))
```

```{r mfa-bootstrap-plot, echo = FALSE, fig.cap = "Bootstrap summary for MFA component strengths. Error bars show percentile intervals over resampled refits."}
ggplot(boot_mfa$summary, aes(component, observed)) +
  geom_point(size = 2.5, color = "#9c2f2f") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.12, color = "#9c2f2f") +
  geom_point(aes(y = mean), shape = 1, size = 2.5, color = "grey35") +
  scale_x_continuous(breaks = boot_mfa$summary$component) +
  labs(x = "Component", y = "Singular value")
```

### Permutation test for shared signal

Permutation inference asks whether the observed leading components are stronger
than what you would expect after breaking row-wise alignment across blocks.

```{r mfa-permutation}
perm_mfa <- infer_muscal(
  fit_mfa,
  method = "permutation",
  statistic = "sdev",
  nrep = 19,
  seed = 202
)

perm_mfa$component_results
```

```{r mfa-permutation-check}
stopifnot(all(perm_mfa$component_results$p_value >= 0))
stopifnot(all(perm_mfa$component_results$p_value <= 1))
```

## Reconstruction cross-validation for MFA

For same-row methods, the usual held-out target is reconstruction error.

```{r mfa-cv-setup}
X_concat <- do.call(cbind, sim$data_list)
md <- multidesign::multidesign(X_concat, data.frame(batch = rep(c("A", "B"), each = 24)))
folds_mfa <- multidesign::cv_rows(
  md,
  rows = list(1:6, 25:30),
  preserve_row_ids = TRUE
)
```

```{r mfa-cv}
res_mfa_cv <- cv_muscal(
  folds = folds_mfa,
  fit_fn = function(analysis) {
    Xa <- multidesign::xdata(analysis)
    mfa(
      list(
        X1 = Xa[, 1:12, drop = FALSE],
        X2 = Xa[, 13:22, drop = FALSE],
        X3 = Xa[, 23:30, drop = FALSE]
      ),
      ncomp = 2
    )
  },
  estimate_fn = function(model, assessment) {
    predict(model, multidesign::xdata(assessment), type = "reconstruction")
  },
  truth_fn = function(assessment) multidesign::xdata(assessment),
  metrics = c("mse", "rmse", "r2")
)

res_mfa_cv$scores
```

```{r mfa-cv-check}
stopifnot(all(is.finite(res_mfa_cv$scores$mse)))
stopifnot(all(res_mfa_cv$scores$rmse >= 0))
```

## Response-prediction cross-validation for Anchored MFA

Anchored MFA has a different evaluation target: predicting rows of the
reference block `Y` from auxiliary blocks with incomplete overlap.

```{r simulate-anchored, include = FALSE}
set.seed(42)
N <- 36
r <- 2

F_true <- matrix(rnorm(N * r), N, r)
F_true <- qr.Q(qr(scale(F_true, scale = FALSE)))

B_ref <- svd(matrix(rnorm(9 * r), 9, r))$u
Y <- F_true %*% t(B_ref) + 0.2 * matrix(rnorm(N * 9), N, 9)
Y <- scale(Y, scale = FALSE)
colnames(Y) <- paste0("y", seq_len(ncol(Y)))

idx1 <- sort(sample.int(N, 24))
idx2 <- sort(sample.int(N, 22))
V1 <- svd(matrix(rnorm(14 * r), 14, r))$u
V2 <- svd(matrix(rnorm(11 * r), 11, r))$u

X1 <- scale(F_true[idx1, ] %*% t(V1) + 0.2 * matrix(rnorm(length(idx1) * 14), length(idx1), 14), scale = FALSE)
X2 <- scale(F_true[idx2, ] %*% t(V2) + 0.2 * matrix(rnorm(length(idx2) * 11), length(idx2), 11), scale = FALSE)
```

```{r anchored-fit}
fit_anchor <- anchored_mfa(
  Y = Y,
  X = list(X1 = X1, X2 = X2),
  row_index = list(X1 = idx1, X2 = idx2),
  ncomp = 2
)

stopifnot(all(is.finite(scores(fit_anchor))))
stopifnot(setequal(fit_anchor$oos_types, c("response", "scores", "reconstruction")))
```

```{r anchored-cv-setup}
d1 <- multidesign::multidesign(X1, data.frame(anchor = idx1, Y[idx1, , drop = FALSE]))
d2 <- multidesign::multidesign(X2, data.frame(anchor = idx2, Y[idx2, , drop = FALSE]))
hd <- multidesign::hyperdesign(list(X1 = d1, X2 = d2), block_names = c("X1", "X2"))

folds_anchor <- multidesign::cv_rows(
  hd,
  rows = list(
    list(X1 = 1:3, X2 = 1:3),
    list(X1 = 4:6, X2 = 4:6)
  ),
  preserve_row_ids = TRUE
)
```

```{r anchored-cv}
res_anchor_cv <- cv_muscal(
  folds = folds_anchor,
  fit_fn = function(analysis) {
    X_blocks <- multidesign::xdata(analysis)
    idx_blocks <- lapply(multidesign::design(analysis), function(des) des$anchor)
    anchored_mfa(Y = Y, X = X_blocks, row_index = idx_blocks, ncomp = 2)
  },
  estimate_fn = function(model, assessment) {
    X_blocks <- multidesign::xdata(assessment)
    do.call(rbind, lapply(names(X_blocks), function(block_name) {
      predict(model, X_blocks[[block_name]], block = block_name, type = "response")
    }))
  },
  truth_fn = function(assessment) {
    des_blocks <- multidesign::design(assessment)
    do.call(rbind, lapply(des_blocks, function(des) {
      as.matrix(des[, colnames(Y), drop = FALSE])
    }))
  },
  task = "response_prediction",
  metrics = c("mse", "mean_cosine_similarity")
)

res_anchor_cv$scores
```

```{r anchored-cv-check}
stopifnot(all(is.finite(res_anchor_cv$scores$mse)))
stopifnot(all(is.finite(res_anchor_cv$scores$mean_cosine_similarity)))
```

## Inference for graph-anchored models

`graph_anchored_mfa()` uses the same anchored prediction target, but adds a
graph penalty over auxiliary features. The evaluation workflow stays the same:
you can still use `infer_muscal()` for resampling summaries and `cv_muscal()`
for held-out prediction.

```{r graph-anchored-fit}
colnames(X1) <- c(paste0("f", 1:4), paste0("u1_", 1:10))
colnames(X2) <- c(paste0("f", 1:4), paste0("u2_", 1:7))

fit_graph <- graph_anchored_mfa(
  Y = Y,
  X = list(X1 = X1, X2 = X2),
  row_index = list(X1 = idx1, X2 = idx2),
  ncomp = 2,
  feature_graph = "colnames",
  graph_lambda = 2
)

stopifnot(fit_graph$fit_spec$refit_supported)
stopifnot(all(is.finite(scores(fit_graph))))
```

```{r graph-anchored-bootstrap}
boot_graph <- infer_muscal(
  fit_graph,
  method = "bootstrap",
  statistic = "sdev",
  nrep = 6,
  seed = 303
)

boot_graph$summary
```

```{r graph-anchored-check}
stopifnot(all(is.finite(boot_graph$summary$mean)))
stopifnot(all(boot_graph$summary$upper >= boot_graph$summary$lower))
```

## Choosing the right validation path

- Use `infer_muscal()` when the fitted method stores refit metadata and you want
  bootstrap intervals or permutation p-values.
- Use `cv_muscal()` when you care about held-out reconstruction or prediction
  error.
- Use method-specific tuning helpers such as `ipca_tune_alpha()` when the model
  has a dedicated hyperparameter-selection routine.

Not every method currently exposes generic refit metadata. When that is not yet
available, you can still use `cv_muscal()` by providing explicit `fit_fn`,
`estimate_fn`, and `truth_fn` callbacks. At present, `mfa()`,
`mcca()`, `ipca()`, `anchored_mfa()`, `aligned_mfa()`, and
`graph_anchored_mfa()` all support generic `infer_muscal()` workflows directly.

## Next steps

- `vignette("mfa")` for the core same-row MFA workflow
- `vignette("mcca")` for correlation-driven same-row integration
- `vignette("ipca")` for adaptive integration with both tuning and generic evaluation
- `vignette("linked_mfa")` for anchored prediction with incomplete overlap
- `?graph_anchored_mfa` for graph-regularized anchored integration
