---
title: Penalized Multiple Factor Analysis
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Penalized Multiple Factor Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE) && exists("params")) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
```

## Why Penalized MFA?

Standard `mfa()` normalizes blocks so they contribute equally, but it places
no constraint on the *loadings* themselves. Each block's loadings can point in
completely different directions. When you expect blocks to share a similar
variable structure — say, the same brain regions measured under different
conditions — you want corresponding variables to load similarly across blocks.

Penalized MFA adds a regularization penalty that encourages loading matrices
to be similar, while still allowing each block to express its unique structure.
The `lambda` parameter controls the trade-off: `lambda = 0` recovers
independent per-block PCA; large `lambda` forces near-identical loadings.

## Quick start

```{r load-packages}
library(muscal)
library(multivarious)
library(ggplot2)
```

```{r simulate-data}
sim <- synthetic_multiblock(
  S = 4, n = 60, p = 25, r = 3,
  sigma = 0.3, seed = 42
)
sapply(sim$data_list, dim)
```

Fit Penalized MFA with a moderate penalty:

```{r fit-pmfa}
fit <- penalized_mfa(
  sim$data_list, ncomp = 3,
  lambda = 1, max_iter = 20
)
fit
```

```{r quick-score-plot, echo = FALSE, fig.cap = "Consensus scores from Penalized MFA. Each observation is positioned by the mean of its block-specific scores."}
autoplot(fit)
```

## How it works

Penalized MFA estimates orthonormal loading matrices
$\mathbf{V}_i \in \mathbb{R}^{p \times k}$ for each block by minimizing:

$$\sum_{i=1}^{S} \|\mathbf{X}_i - \mathbf{X}_i \mathbf{V}_i \mathbf{V}_i^\top\|_F^2
\;+\; \lambda \cdot \text{Penalty}(\{\mathbf{V}_i\})$$

The first term is reconstruction error; the second encourages similarity among
block loadings. The optimization uses Riemannian gradient descent on the
Stiefel manifold (the space of orthonormal matrices).

## Convergence

```{r convergence, fig.cap = "Objective value over iterations. The penalty drives blocks toward shared loading structure while the reconstruction term keeps them data-faithful."}
plot_convergence(fit)
```

## Visualization

### Consensus scores

The consensus score is the mean of per-block scores. You can also view
individual block scores:

```{r autoplot-block, fig.cap = "Scores from block 1 alone."}
autoplot(fit, block = 1)
```

### Variance explained

```{r variance, fig.cap = "Variance explained by each consensus component."}
plot_variance(fit, type = "bar")
```

### Partial factor scores

Compare how blocks position the same observations:

```{r partial-scores, fig.cap = "Partial factor scores. Each color is a block; lines connect to the consensus. The penalty shrinks these distances compared to standard MFA.", fig.width = 7, fig.height = 5}
plot_partial_scores(fit, connect = TRUE, show_consensus = TRUE)
```

### Variable loadings

```{r loadings, fig.cap = "Top 15 variable loadings on component 1, colored by block."}
plot_loadings(fit, type = "bar", component = 1, top_n = 15)
```

## Penalty methods

Three penalty formulations are available:

- **`"projection"`** (default): Penalizes the distance between block-specific
  projection matrices $\mathbf{P}_i = \mathbf{V}_i \mathbf{V}_i^\top$.
  Rotation-invariant — only the subspace matters, not the basis.

- **`"pairwise"`**: Penalizes all pairwise distances between loading matrices
  directly. Sensitive to rotation.

- **`"global_mean"`**: Penalizes each block's deviation from the mean loading
  matrix.

```{r penalty-methods, eval = FALSE}
# Projection penalty (default, recommended)
fit_proj <- penalized_mfa(sim$data_list, ncomp = 3,
                          lambda = 1, penalty_method = "projection")

# Pairwise penalty
fit_pair <- penalized_mfa(sim$data_list, ncomp = 3,
                          lambda = 1, penalty_method = "pairwise")
```

## Choosing lambda

The penalty strength `lambda` is the key tuning parameter:

- **`lambda = 0`**: No penalty. Each block gets its own independent PCA
  loadings.
- **Small `lambda`** (0.01–0.1): Gentle encouragement toward similar loadings.
- **Moderate `lambda`** (1–10): Strong similarity constraint while still
  allowing block-specific variation.
- **Large `lambda`** (100+): Forces near-identical loadings across blocks.

A practical approach is to compare reconstruction error across a grid of lambda
values and pick the knee of the curve:

```{r lambda-grid, eval = FALSE}
lambdas <- c(0.01, 0.1, 1, 10, 100)
fits <- lapply(lambdas, function(lam) {
  penalized_mfa(sim$data_list, ncomp = 3, lambda = lam, max_iter = 20)
})
```

## Optimizer options

By default, Penalized MFA uses the Adam optimizer on the Stiefel manifold.
You can also use plain gradient descent:

```{r optimizer, eval = FALSE}
fit_gd <- penalized_mfa(
  sim$data_list, ncomp = 3, lambda = 1,
  optimizer = "gradient",
  learning_rate = 0.001,
  max_iter = 50
)
```

## When to use Penalized MFA

Penalized MFA is appropriate when:

- You have multiple blocks with the same variables (or matched variables)
- You expect blocks to share similar loading structure
- You want to control the similarity-reconstruction trade-off explicitly
- Standard MFA loadings are too noisy or divergent across blocks

Penalized MFA is *not* appropriate when:

- Blocks measure entirely different variables (the loading penalty is
  meaningless)
- You need classical MFA normalization guarantees (use `mfa()`)
- Your goal is block-level outlier detection (use `covstatis()`)

## Next steps

- `vignette("mfa")` — Standard MFA without loading penalties
- `vignette("ipca")` — iPCA, an alternative integration approach
- `?penalized_mfa_clusterwise` — Cluster-wise penalized MFA

## References

The optimization is based on block-coordinate descent with Riemannian updates
on the Stiefel manifold, using either Adam or plain gradient descent with
retraction via QR decomposition.
