---
title: Choosing a Supervised Common-Space Model
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a Supervised Common-Space Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE) && exists("params")) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
```

## Why this comparison matters

Sometimes the response table really defines a small set of shared states across
blocks. Sometimes every block has its own paired multivariate response with no
exact row sharing. Sometimes you are in between: some rows map back to known
anchor states, while others are novel.

`muscal` now has three honest answers to those three situations:

- `anchored_mfa()` when the response lives at the anchor-state level
- `response_aligned_mfa()` when each block has paired multivariate `Y`, with or
  without partial anchor structure
- `aligned_rrr()` when you want the clean reduced-rank regression baseline

This vignette shows the same prediction problem under each regime and focuses
on the contract you actually need, not on forcing every case through repeated
rows of `Y`.

## Load the package

```{r load-packages}
library(muscal)
library(multivarious)
library(knitr)
```

```{r simulate-data, include = FALSE}
.center_only <- function(x) scale(x, center = TRUE, scale = FALSE)

.mean_mse <- function(pred, truth) {
  mean(unlist(lapply(seq_along(pred), function(k) {
    (pred[[k]] - truth[[k]])^2
  })))
}

.predict_block_list <- function(fit, X_list, type = c("response", "scores"), ...) {
  type <- match.arg(type)
  lapply(names(X_list), function(block) {
    predict(fit, X_list[[block]], block = block, type = type, ...)
  }) |> stats::setNames(names(X_list))
}

.sim_exact_anchor <- function(seed = 101,
                              n_anchor = 24,
                              n_train = c(54, 50),
                              n_test = c(18, 16),
                              p = c(8, 7),
                              q = 3,
                              K = 2,
                              noise_x = 0.07,
                              noise_y = 0.04) {
  set.seed(seed)
  S_anchor <- .center_only(matrix(rnorm(n_anchor * K), n_anchor, K))
  B <- matrix(rnorm(q * K), q, K)
  V_list <- lapply(p, function(pk) matrix(rnorm(pk * K), pk, K))

  Y_anchor <- .center_only(S_anchor %*% t(B) + matrix(rnorm(n_anchor * q, sd = noise_y), n_anchor, q))

  make_split <- function(n_obs) {
    idx <- lapply(n_obs, function(nk) sample.int(n_anchor, nk, replace = TRUE))
    names(idx) <- c("assay1", "assay2")
    X <- list(
      assay1 = .center_only(S_anchor[idx$assay1, , drop = FALSE] %*% t(V_list[[1]]) +
        matrix(rnorm(length(idx$assay1) * p[[1]], sd = noise_x), length(idx$assay1), p[[1]])),
      assay2 = .center_only(S_anchor[idx$assay2, , drop = FALSE] %*% t(V_list[[2]]) +
        matrix(rnorm(length(idx$assay2) * p[[2]], sd = noise_x), length(idx$assay2), p[[2]]))
    )
    truth <- list(
      assay1 = Y_anchor[idx$assay1, , drop = FALSE],
      assay2 = Y_anchor[idx$assay2, , drop = FALSE]
    )
    list(X = X, row_index = idx, Y_truth = truth)
  }

  list(
    Y_anchor = Y_anchor,
    train = make_split(n_train),
    test = make_split(n_test)
  )
}

.sim_blockwise_response <- function(seed = 202,
                                    n_anchor = 22,
                                    n_train = c(48, 44),
                                    n_test = c(16, 14),
                                    p = c(8, 7),
                                    q = 3,
                                    K = 2,
                                    noise_x = 0.12,
                                    noise_y = 0.05,
                                    latent_jitter = 0.12,
                                    missing_anchor_frac = 0.35) {
  set.seed(seed)
  S_anchor <- .center_only(matrix(rnorm(n_anchor * K), n_anchor, K))
  B <- matrix(rnorm(q * K), q, K)
  V_list <- lapply(p, function(pk) matrix(rnorm(pk * K), pk, K))
  anchor_response <- .center_only(S_anchor %*% t(B) + matrix(rnorm(n_anchor * q, sd = noise_y), n_anchor, q))

  make_split <- function(n_obs) {
    idx <- lapply(n_obs, function(nk) sample.int(n_anchor, nk, replace = TRUE))
    names(idx) <- c("assay1", "assay2")
    Z <- list(
      assay1 = S_anchor[idx$assay1, , drop = FALSE] +
        matrix(rnorm(length(idx$assay1) * K, sd = latent_jitter), length(idx$assay1), K),
      assay2 = S_anchor[idx$assay2, , drop = FALSE] +
        matrix(rnorm(length(idx$assay2) * K, sd = latent_jitter), length(idx$assay2), K)
    )
    X <- list(
      assay1 = .center_only(Z$assay1 %*% t(V_list[[1]]) +
        matrix(rnorm(length(idx$assay1) * p[[1]], sd = noise_x), length(idx$assay1), p[[1]])),
      assay2 = .center_only(Z$assay2 %*% t(V_list[[2]]) +
        matrix(rnorm(length(idx$assay2) * p[[2]], sd = noise_x), length(idx$assay2), p[[2]]))
    )
    Y <- list(
      assay1 = .center_only(Z$assay1 %*% t(B) +
        matrix(rnorm(length(idx$assay1) * q, sd = noise_y), length(idx$assay1), q)),
      assay2 = .center_only(Z$assay2 %*% t(B) +
        matrix(rnorm(length(idx$assay2) * q, sd = noise_y), length(idx$assay2), q))
    )
    anchor_map <- lapply(idx, function(ii) {
      out <- ii
      miss <- sample.int(length(ii), size = ceiling(missing_anchor_frac * length(ii)))
      out[miss] <- NA_integer_
      out
    })
    list(X = X, Y = Y, anchor_map = anchor_map, full_anchor_map = idx)
  }

  list(
    anchor_response = anchor_response,
    train = make_split(n_train),
    test = make_split(n_test)
  )
}

exact_anchor <- .sim_exact_anchor()
blockwise <- .sim_blockwise_response()
```

## Quick start: paired responses, no exact anchor assumption

If every block has its own paired multivariate response, start with
`response_aligned_mfa()`. If you want the cleaner supervised baseline with no
predictor reconstruction term, fit `aligned_rrr()` alongside it.

```{r quick-start-fits}
fit_response <- response_aligned_mfa(
  Y = blockwise$train$Y,
  X = blockwise$train$X,
  ncomp = 2,
  preproc = multivarious::pass(),
  response_preproc = multivarious::pass(),
  normalization = "None",
  ridge = 1e-8,
  max_iter = 80,
  tol = 1e-9
)

fit_rrr <- aligned_rrr(
  Y = blockwise$train$Y,
  X = blockwise$train$X,
  ncomp = 2,
  preproc = multivarious::pass(),
  response_preproc = multivarious::pass(),
  ridge = 1e-8,
  max_iter = 80,
  tol = 1e-9
)
```

```{r quick-start-checks, include = FALSE}
pred_response <- .predict_block_list(fit_response, blockwise$test$X, type = "response")
pred_rrr <- .predict_block_list(fit_rrr, blockwise$test$X, type = "response")

quick_summary <- data.frame(
  model = c("response_aligned_mfa()", "aligned_rrr()"),
  mean_test_mse = c(
    .mean_mse(pred_response, blockwise$test$Y),
    .mean_mse(pred_rrr, blockwise$test$Y)
  )
) |>
  transform(mean_test_mse = round(mean_test_mse, 4))

stopifnot(all(is.finite(quick_summary$mean_test_mse)))
stopifnot(all(quick_summary$mean_test_mse > 0))
stopifnot(fit_response$score_representation == "stacked_block_scores")
stopifnot(fit_rrr$score_representation == "stacked_block_scores")
```

```{r quick-start-summary}
kable(quick_summary, align = c("l", "r"))
```

Both models predict from block-specific `X` into a shared multivariate response
space. The practical difference is structural: `response_aligned_mfa()` also
models the predictor blocks through a shared latent space, while
`aligned_rrr()` is the lean reduced-rank baseline.

## Which data contract do you actually have?

```{r regime-table}
chooser <- data.frame(
  Data_regime = c(
    "Exact shared anchor states",
    "Some anchored rows, some novel rows",
    "No anchor states, paired blockwise responses"
  ),
  What_is_observed = c(
    "One anchor-level Y table plus row maps from each X block",
    "Blockwise Y_k plus optional anchor_map for the rows you can link",
    "Only blockwise paired (X_k, Y_k)"
  ),
  Recommended_fit = c(
    "anchored_mfa()",
    "response_aligned_mfa()",
    "response_aligned_mfa() plus aligned_rrr() baseline"
  )
)

kable(chooser, align = "l")
```

The model choice should follow that contract directly. The point is not to
force every problem into repeated rows of `Y`, but to keep the common-space
assumption honest.

## When the response really is an anchor table

Use `anchored_mfa()` when the scientifically meaningful response lives at the
anchor-state level and each predictor block only tells you which anchor rows it
touches.

```{r anchored-fit}
fit_anchor <- anchored_mfa(
  Y = exact_anchor$Y_anchor,
  X = exact_anchor$train$X,
  row_index = exact_anchor$train$row_index,
  ncomp = 2,
  preproc = multivarious::pass(),
  normalization = "None",
  ridge = 1e-8,
  max_iter = 80,
  tol = 1e-9
)
```

```{r anchored-checks, include = FALSE}
pred_anchor <- .predict_block_list(fit_anchor, exact_anchor$test$X, type = "response")
anchor_summary <- data.frame(
  model = "anchored_mfa()",
  mean_test_mse = round(.mean_mse(pred_anchor, exact_anchor$test$Y_truth), 4),
  n_anchor_states = nrow(exact_anchor$Y_anchor),
  learned_score_rows = nrow(multivarious::scores(fit_anchor))
)

stopifnot(all(is.finite(anchor_summary$mean_test_mse)))
stopifnot(anchor_summary$learned_score_rows == anchor_summary$n_anchor_states)
```

```{r anchored-summary}
kable(anchor_summary, align = c("l", "r", "r", "r"))
```

That is the clean anchored contract: the shared score table lives over anchor
states, not over block rows, and new `X` rows are projected back to that
anchor-level response surface.

## When some rows are anchored but others are novel

This is the regime `response_aligned_mfa()` was generalized to handle. You keep
the blockwise responses `Y_k`, pass anchor information only where it is real,
and let the novel rows stay free.

```{r hybrid-fit}
fit_hybrid <- response_aligned_mfa(
  Y = blockwise$train$Y,
  X = blockwise$train$X,
  ncomp = 2,
  preproc = multivarious::pass(),
  response_preproc = multivarious::pass(),
  normalization = "None",
  anchor_response = blockwise$anchor_response,
  anchor_map = blockwise$train$anchor_map,
  coupling_lambda = 2,
  ridge = 1e-8,
  max_iter = 80,
  tol = 1e-9
)
```

```{r hybrid-checks, include = FALSE}
pred_hybrid_plain <- .predict_block_list(
  fit_hybrid,
  blockwise$test$X,
  type = "response"
)
pred_hybrid_anchor <- lapply(names(blockwise$test$X), function(block) {
  predict(
    fit_hybrid,
    blockwise$test$X[[block]],
    block = block,
    type = "response",
    new_anchor_map = blockwise$test$full_anchor_map[[block]]
  )
}) |> stats::setNames(names(blockwise$test$X))

hybrid_summary <- data.frame(
  prediction_path = c("X only", "X + test-time anchor_map"),
  mean_test_mse = c(
    .mean_mse(pred_hybrid_plain, blockwise$test$Y),
    .mean_mse(pred_hybrid_anchor, blockwise$test$Y)
  )
) |>
  transform(mean_test_mse = round(mean_test_mse, 4))

stopifnot(all(is.finite(hybrid_summary$mean_test_mse)))
stopifnot(is.matrix(fit_hybrid$S))
stopifnot(nrow(fit_hybrid$S) == nrow(blockwise$anchor_response))
```

```{r hybrid-summary}
kable(hybrid_summary, align = c("l", "r"))
```

The important point is not just that the fit stores both `Z_list` and `S`. It
is that prediction stays honest: default prediction uses only `X`, while known
test-time anchor information can refine the score solve when you truly have it.

## How should you choose?

- Use `anchored_mfa()` when the shared response really is an anchor-state
  object and your blocks connect to it through row maps.
- Use `response_aligned_mfa()` when each block has its own paired multivariate
  response and you want a shared latent geometry for the predictor blocks.
- Add `anchor_map` and `anchor_response` to `response_aligned_mfa()` only when
  those anchors are genuine; missing anchor rows should stay missing.
- Fit `aligned_rrr()` alongside `response_aligned_mfa()` when prediction is the
  main target and you want to know whether the extra predictor-reconstruction
  structure is actually buying anything.

## Where next?

- `vignette("linked_mfa")` for the anchored row-mapping workflow in more depth
- `vignette("aligned_mfa")` for the unsupervised same-latent-row analogue
- `vignette("model_evaluation")` for bootstrap, permutation, and held-out
  evaluation workflows across supported methods
- `?response_aligned_mfa` and `?aligned_rrr` for the full argument contracts
